How to publish in Biological Sciences: a guide for the uninitiated

### 2021-06-10 4.01-openaccess.Rmd
Add files via upload
# UPLOAD working version that prints to pdf and gitbook

## So what is open access?
I have covered the many different kinds of Open Access (OA) elsewhere in this book ([see part 4](openaccess1.html)). Here I concentrate on the fiscal implication of OA. Someone needs to pay for the work done. Who should pay and how?

Open access is probably one of the best scams that the publishers have come up with to date. Now the scientists pay for making their own content open for anyone to read. They pay a once off fee to the publishers to typeset the manuscript and host it on their site without a paywall. And how much do the publishers want for this service. Prices start from USD 1000 and go up to around USD 12 000. More frightening, and evidence that publishers are simply taking advantage of scientists, is that prices increase with the Impact Factor of the journal [@gray2020sorry], although the costs involved to the publisher remain static.

The money comes from the funds that would otherwise be reserved for conducting science. So now the money for research goes directly into the pockets of the publishers upfront. Money that ultimately comes from you as tax-payers goes directly to publishers. Still happy?

## So does that mean that these journals are now free?
Mostly no. The majority OA model ([hybrid OA](openaccess1.html)) means that a minority of articles in these journals are free, but the universities are expected to subscribe to those same journals at ever increasing prices because much of the rest of the content is still behind the [paywall](paywall.html). This is because most authors cannot afford to pay the exorbitant fees charged by the journals (although some countries now have this payment as mandatory - PlanS, they and their scientists are still in a minority). There are some journals that are entirely open access ([gold OA](openaccess1.html)). These are (almost) exclusively online and have never been part of traditional packages that university libraries spend so much of their budget on. Hence the fact that they are entirely free does not impact library budgets. As a scientist, it will make an overall increase to papers that you can access.

@@

@@

For a lot of academics there is a trade-off: vanity for cash. We know that for a lot of academics, their vanity will mean that they are prepared to find the money (even if it comes from their own pockets). However, for as long as enough are willing to pay, we will continue to be exploited. Instead, we really need a new model for publishing. For some academics they will be paying directly for a chance at [getting a job](jobinacademia.html) or [getting tenure](habilitation.html). The chances of these individuals reaching their goal from a privaledged institution then simply becomes reinforced.

@@

## A new model for publishing without publishers - the blockchain journal

### The problem
There are growing problems with publishers. Publishing has become very expensive for scientists, and the public who fund science. Many funders are now unwittingly funding publishers instead of science. I’ve discussed the history of scientific publishing before (here), and explained how commercial publishers have had a role right from the very beginning. This relationship continues to the present day, but while the origins saw learned societies commissioning publishers and then distributing their content, today the publishers have climbed into the driving seat, conceiving and owning many of the current journal titles. This is not to say that there are no scholarly societies that still dictate terms to publishers (The British Ecological Society, BES, do this very well), but they are few and far between, and even when society journals are involved, the goliath publishing companies easily outweigh any control that they might have once possessed (and I’m speaking from personal experience - [see part 5](paywall.html)).

The result is that publishers have become gigantic corporations that now dictate to the scientists that produce, edit and review all of the content. They charge incredibly high fees to anyone who might want to read the publicly funded content. The budgets of university libraries run into USD 100 000s just to access content. The result is that many universities can’t afford all of the content that their researchers need. Publicly funded content, and by that I mean that you dear reader are paying for the original science of the content in your taxes (yes, you all pay taxes, even if it’s only VAT), and then you pay a second time for the researchers (who themselves produced the content) to access the content. Who benefits from the fact that you pay twice? The publishers. Why aren’t you upset about this? Probably because you are unaware. But if you are upset, then join in the discussion to decide how to emancipate ourselves from the publishers who are merrily munching on through this publicly funded cash cow.

### What do publishers do?
The publishers would claim that they do an awful lot. All they really do is pay for the layout and printing of journals. These days ‘printing’ really means hosting electronic pdfs only, as there’s very little paper that’s printed, and you can be sure that paper subscribers now pay the extra cost of any additional fees. The layout from the manuscript (most often a MS Word doc) into a pdf does take some skill and talent, although nothing like what you might expect given how much the publishers charge. You can be sure that they don’t pay much for this service as almost all layout is done in India, Bangladesh, Sri Lanka, etc. Quality can be good, but more often quality control is completely lacking. Most authors have stories of how manuscripts have come back mangled, although my own impression is that the worst days appear to be over.

This is not to say that there are no skills in the publishing world. I also want to make it clear that I am not condemning all those employed by all publishing houses. I have interacted with many excellent staff at publishing houses that have some of the worst practices. It’s important to note that these staff do not get the fiscal benefit of the aggressive and unethical behaviour of their publishing companies. These profits are retained for the directors and shareholders. One might also question whether the sales reps of these companies who make deals with academic libraries don’t also receive unethical bonuses for their undisclosed financial arrangements.

@@

To get behind that paywall, university libraries need to subscribe. Publishers bundle journals together and sell subscriptions at very high prices. If you are inside the university IP address, this access should be seamless. If you are outside, you might need to log in through your university’s library. There are other tactics for getting around a paywall ([see part 2](paywall.html)).

So far, the publisher hasn’t produced any content. The scientific content has been produced by the scientist at the cost of the public purse. The editing and peer review ([see part 4](preprint.html)) has all been done by the scientists, which has also cost the public purse but has been completely free to the publisher. OK, so there are some small costs associated with manuscript handling software subscriptions that the publishers normally pay. The publisher has also paid for the typesetting (although they’ve done this as cheaply as possible - see above), and they’ve paid for the servers that distribute the pdfs maintaining that all important paywall. What else? Nothing else. Now they simply charge everyone to look at the content (and because it’s by subscription), actually charge everyone whether or not they are looking at the content. Bergstrom and Bergstrom [-@bergstrom2006economics] showed the difference between the costs of subscribing to society journals in 2005 was US$ 0.29 while for profit journals cost US$ 1.42, irrespective of any degree of quality. Fifteen years on, most society journals have now been captured by for profit publishers and now we all pay the higher price.

@@

>>>>

## So what is open access?
I have covered the many different kinds of Open Access (OA) elsewhere in this book ([see part 2](openaccess1.html)). Here I concentrate on the fiscal implication of OA. Someone needs to pay for the work done. Who should pay and how?

The principle of Open Access is something that is easy to appreciate, and lies very much at the heart at the movement from [closed to open science](#openscience). Some history is appropriate when we talk about Open Access, because the movement started as a genuine attempt to change the publishing model for the better. Hopefully in the future this experiment will be seen as a success. In the meantime, Open Access has been hijacked by publishers as a way in which they can earn large amounts of additional income.

### _PLOS ONE_ {#PLOS}
The Public Library of Science (PLOS) started soon after the turn of the new millennium primarily as a concern to the problems with peer review that are highlighted throughout this book. This really was a new experiment in publishing, that was to radically alter the publishing model in the 21st century, and in only 10 years take the place as the world's biggest journal. The idea was to do away with all the normal waiting times in publishing, and to have articles published immediately online, and without any charges associated with accessing them: Open Access. At the time, this was all fairly radical, and it became very successful with a large amount of interest from scientists who submitted their articles to PLOS. But the founders were frustrated that they were rejecting a lot of papers that were technically good, but not selected as they were not novel, or did not signicantly advance their field. In 2006, PLOS dreamt up a new journal to take all of these technically sound manuscripts, and thus was the birth of _PLOS ONE_.

From the outset, _PLOS ONE_ was not interested in acquiring any [Impact Factor](impactfactor.html). Instead, it was mostly interested in taking technically sound manuscripts irrespective of their results. This meant that reviewers for _PLOS ONE_ were (and still are) asked to assess the technical soundness of a manuscript, and not to judge the 'value' of the results. The original idea was that reviews should be post publication, through comments made on the website after the publications were posted. The initial review then, was simply meant as a preliminary check for technical soundness. This was again a radical departure from the publishing norm by the PLOS group, when many other journals were still [rejecting](#rejection) manuscripts that had no faults other than they were not attractive to editors. _PLOS ONE_ was opened to accept all such manuscripts, effectively doing away with [publication bias](#badscience). As you might imagine, at a time when academics were under pressure to ['publish or perish'](publishmore.html) rejections were plentiful and so manuscripts quickly found their way to _PLOS ONE_. Because _PLOS ONE_ was also Open Access, it's papers received an extra boost of citations through visibility, and ironically after 5 years it found that it had an Impact Factor to rival some of the journals whose rejections it was picking up. This meant that more and more scientists started submitting to _PLOS ONE_ as their publication of choice, mostly because it was publishing Open Access.

_PLOS ONE_ became the worlds biggest scientific journal, and in 2013 it published 32058 papers [@davies2019plos take a moment to do the math on the income that PLOS received for these at > USD 1000 per article]. By this time, other academic publishers had noticed the rocketing ascent of _PLOS ONE_ and had responded by starting their own 'no impact factor' journals that could be fed by the rejection piles of journals already in their stables. Notable among these was _Scientific Reports_ from the Nature Publishing Group which started in 2011 with 208 papers, and after 5 years ramping up to nearly 20000 papers in 2018. For whatever reason, the Impact Factor of _Scientific Reports_ is roughly double that of PLOS ONE for around the same APC, drawing far more researchers in. This same period, since 2013, has seen many such examples of 'no impact factor' journals that accept any article that is technically sound. They are not only the general behemoths like _PLOS ONE_ and _Scientific Reports_, but you will find that even niche journals that make significant numbers of rejections of technically sound papers have their own version that they will offer to you in the same email that informs you of a [rejection](#rejection) from their flagship journal.

The bottom line, from this story of _PLOS ONE_, the highly innovative journal with an attempt to fundamentally change the publishing landscape, is that this is exactly what it did. However, somewhere along the way the well meaning ideas were hijacked by mainstream publishers' greed. They saw a massive and untapped market. Unsurprisingly, they decided not to call it 'no impact factor' journals, or even the 'rejection pile', but they named it [Open Access](openaccess.html).

The bandwagon movement to Open Access that happened as a result of _PLOS ONE_ success, is probably one of the best scams that the publishers have come up with to date. Now the scientists pay for making their own content open for anyone to read. They pay a once off fee to the publishers to typeset the manuscript and host it on their site without a paywall. And how much do the publishers want for this service. Prices start from USD 1000 and go up to around USD 12 000. More frightening, and evidence that publishers are simply taking advantage of scientists, is that prices increase with the [Impact Factor](impactfactor.html) of the journal [@gray2020sorry], although the costs involved to the publisher remain static. What further evidence do you need?

## So does that mean that OA journals are now free?
Mostly no. The majority OA model ([hybrid OA](openaccess1.html)) means that a minority of articles in these journals are free, but the universities are expected to subscribe to those same journals at ever increasing prices because much of the rest of the content is still behind the [paywall](paywall.html). This is because most authors cannot afford to pay the exorbitant fees charged by the journals (although some countries now have this payment as mandatory - [cOAlition S](https://www.coalition-s.org/), they and their scientists are still in a minority). There are some journals that are entirely open access ([gold OA](#gold) and [Diamond OA](#diamond)). These are (almost) exclusively online and have never been part of traditional packages that university libraries spend so much of their budget on. Hence the fact that they are entirely free does not impact library budgets. The new way to fleece our libraries is through '[transformative publishing agreements](paywall.html)' As a scientist, it will make an overall increase to papers that you can access.

@@

There are people that are trying to stop this constant escalation of publishers wanting money. You as an Early Career Researcher are in one of the best positions to do something about this change from [closed to open science](#openscience). Currently, [Diamond OA](#diamond) journals are a very small and quite unusual component of the publishing scene. But it is totally possible for these to become the mainstream using initiatives like the [overlay](#overlay) system. The current dynamic will only change when academics submitting their manuscripts change from for-profit publishers to other models. The power is literally in your choice of [where to send your manuscript](whichjournal.html).

@@

For a lot of academics there is a trade-off: vanity for cash. We know that for a lot of academics, their vanity will mean that they are prepared to find the money (even if it comes from their own pockets). However, for as long as enough are willing to pay, we will continue to be exploited. Instead, we really need a new model for publishing. For some academics they will be paying directly for a chance at [getting a job](jobinacademia.html) or [getting tenure](habilitation.html). The chances of these individuals reaching their goal from a privileged institution then simply becomes reinforced.

@@

**We need a new model for publishing without for-profit publishers**

## The problem
There are growing problems with publishers. Publishing has become very expensive for scientists, and the public who fund science. Many funders are now unwittingly funding publishers instead of science. The history of scientific publishing includes a role for commercial publishers right from the very beginning([Part IV](paywall2.html)). This relationship continues to the present day, but while the origins saw learned societies commissioning publishers and then distributing their content, today the publishers have climbed into the driving seat, conceiving and owning many of the current journal titles. Driving the hiring and promotion of scientists that feed their own products ([Part IV](#badscience)), ultimately leading to [fraud and misconduct](#pruittdata). We, as scientists, have allowed this to happen and we are certainly the people that continue to perpetuate the dominance of for-profit publishers in our scientific domain.

As scientists, we need to go back to our core concerns: **the scientific method, independence, transparency and reproducibility**. We need to persuade our institutions that our core concerns are what we should be measured by, and that the new range of Open Science tools be used to determine the extent to which we live up to our scientific values. Getting jobs, tenure and grants should be based on our Open science credentials instead of the publishing industry serving [Impact Factor](impactfactor.html) and other [publishing metrics](citations.html).

### What do publishers do?
The publishers would claim that they do an awful lot. All they really do is pay for the layout and printing of journals. These days ‘printing’ really means hosting electronic pdfs only, as there’s very little paper that’s printed, and you can be sure that paper subscribers now pay the extra cost of any additional fees. The layout from the manuscript (most often a MS Word doc) into a pdf does take some skill and talent, although nothing like what you might expect given how much the publishers charge. You can be sure that they don’t pay much for this service as it is almost all outsourced ([cOAlition S](https://www.coalition-s.org/plan_s_principles/) have insisted that OA journals become transparent with their costs). Quality can be good, but more often quality control is completely lacking. Most authors have stories of how manuscripts have come back mangled, although my own impression is that the worst days appear to be over.

This is not to say that there are no skills in the publishing world. But the reality is that these days it is possible to use free software like Rmarkdown [@xie2018r] to write papers that can quickly and easily be made into any sophisticated layout using LaTex, the same language used by the publishers. Many journals allow submission of articles already formatted.

I also want to make it clear that I am not condemning all those employed by all publishing houses. I have interacted with many excellent staff at publishing houses that have some of the worst practices. It’s important to note that these staff do not get the fiscal benefit of the aggressive and unethical behaviour of their publishing companies. These profits are retained for the directors and shareholders. One might also question whether the sales reps of these companies who make deals with academic libraries don’t also receive unethical bonuses for their undisclosed financial arrangements. But as these deals are not disclosed, who would know or admit this?

@@

To get behind that paywall, university libraries need to subscribe. Publishers bundle journals together and sell subscriptions at very high prices. If you are inside the university IP address, this access should be seamless. If you are outside, you might need to log in through your university’s library. There are other tactics for getting around a paywall ([see part IV](paywall.html)).

So far, the publisher hasn’t produced any content. The scientific content has been produced by the scientist at the cost of the public purse. The editing and peer review ([see part 4](preprint.html)) has all been done by the scientists, which has also cost the public purse but has been completely free to the publisher. OK, so there are some small costs associated with manuscript handling software subscriptions that the publishers normally pay. The publisher has also paid for the typesetting (although they’ve done this as cheaply as possible - see above), and they’ve paid for the servers that distribute the pdfs maintaining that all important paywall. What else? Nothing else. Now they simply charge everyone to look at the content (and because it’s by subscription), actually charge everyone whether or not they are looking at the content. Bergstrom and Bergstrom [-@bergstrom2006economics] showed the difference between the costs of subscribing to society journals in 2005 was USD 0.29 while for profit journals cost US$ 1.42, irrespective of any degree of quality. Fifteen years on, most society journals have now been captured by for profit publishers and now we all pay the higher price.

@@

## Making the change {#change}

My message, throughout this book, has been that the tools to make the shift from [closed to open science](#openscience) are available to us now. There is nothing to stop us changing the system other than the need to act collectively.

[Scholarly societies](#societies) offer an excellent opportunity to organise. Societies still dictate terms to publishers (for example [The British Ecological Society](https://www.britishecologicalsociety.org/), and they have the power to finish their contracts and move their journals to new [Diamond Access](#diamond) formats, using tools such as [overlay journals](#overlay). In turn they have to give up the income that their for-profit publishers supply. This does not mean that they would lose all income, they would still have subscription funds and conferences. But there would be a loss of some of the comfort that they have become used to, making giving up hard to do. But societies need to carry out the wish of their members, and I would suggest that there are more members that would benefit from [Diamond Access](#diamond) society journals, than currently see any benefit in the kick-back from for-profit publishers.

### 2021-06-10 4.03-problempeerreview.Rmd
Add files via upload
# UPLOAD working version that prints to pdf and gitbook

@@

@@

Demonstrated biases in peer review include the following:

- Evidence of bias against female authors is well established [e.g. @tregenza2002gender; @manlove2018authors], and against female reviewers [@helmer2017gender]
- Evidence of bias towards author reputation, favoring acceptance of manuscripts with poor reviews [@bravo2018hidden]
- Evidence of bias towards authors from more prestigious institutions [@tomkins2017single; @manlove2018authors], so-called prestige bias [@lee2013bias]
- Evidence for both nationality and language bias also exist in peer review [@lee2013bias; @manlove2018authors; @nunez2021monolingual]
- Confirmation bias (the tendency for journals and reviewers to favour significant results) is one of the biggest issues for our current system [see @fanelli2010positive and [part IV](transparency.html)]

But that’s not all, as this is only a set of biases that have already been investigated. Given that over 280 biases have already been catalogued (I encourage you to look through the [online catalogue](www.catalogofbias.org), many more different types of bias are likely to exist in peer review. Let’s not forget that our biases have evolved because they are very useful. They exist as a way of shortcutting exhaustive decision making based on random variables. But maybe peer review needs some more of this. And perhaps that means that I should be tolerant when I’m asked to review an economics journal, as these folk clearly weren’t exhibiting any biases associated with economists when they picked me (see [below](editors.html)).

Perhaps the biggest problem facing those who wish to reform the peer review system is that it all starts with editors who are choosing reviewers. Those editors themselves have their own inherent biases. When they look for reviewers, they are likely to sample from within their own group of peers who have the same biases. Interestingly, bias (in general) is more easily perceived by early career scientists [@zvereva2021biases]. My experience is that soliciting reviews from people that I don’t know and have no connection with (are outside of my field) are more likely to fail - they will say no, or they won’t reply to the request [@perry2012peer]. This is even for academics that are publishing within the same area.

@@

## Decisions rest with editors
A good editor will look at the reasoning in the reviews and make a decision in an unbiased way. A poor editor may be swayed by the perceived influence of an important reviewer irrespective of their argument. An increasing trend that I’ve noticed is that editors will simply take a decision that follows the consensus of all reviewers: that is, they rate all reviewers equally (see also Rothwell and Martyn 2000). However, I would argue that this is also bad editing. Irrespective of the bias from reviewers, guarding the integrity of the process of peer review lies with editors.

@@

Casnici et al [-@casnici2017assessing] tracked the fate of rejected manuscripts and showed that if the reviewers had several rounds of peer review before rejection, these manuscripts benefitted later by being accepted to journals with a higher Impact Factor, and/or obtaining greater numbers of citations, even if the reviewers were instrumental in rejecting the manuscript. This suggests that in working collaboratively on a manuscript, reviewers are more likely to promote, cite and help authors. The alternative is that reviewers agree to re-review an article again because they see merit in it, even if they also see flaws. And having spent considerable effort on manuscript, they are more likely to remember and cite it. But this doesn’t take away from the idea that reviewers and authors are collaborating in a social way.

>>>>

There is already a lot covering peer review in this book, and I have placed this chapter last not because it is the least significant, potentially it is the most significant, but because I think that it is important that you appreciate exactly what peer review is, and to some extent experience it, before you begin to consider the problems with the peer review system.

At the heart of the problems with peer review is that individual humans are themselves biased. Because peer review relies on a small number of individuals providing their assessment of a manuscript, it is quite likely that these biases might align, and that the manuscript is rejected along those lines, rather than being considered along purely objective lines. This likelihood of aligned prejudices comes about because the pool of people that conduct peer review in biological sciences, and in many other disciplines, is mostly white, western (i.e. Europe and North America) and male. These people hold a very similar cultural set of biases.

@@

If peer review was effective, then reviews of grant applications should closely align with the productivity of grants given. Fang et al. [-@fang2016nih] found that percentile scores awarded by peer review of NIH grant applications were poor at predicting the the productivity of >100 000 grants awarded.

@@

## Upsetting comments

I think that the reason why we find the comments so harsh is usually because we put so much effort into the writing process that it feels very personal whenever we receive criticism. Indeed, I think that there might be a correlation between how much effort you put in and how harsh the reviewers’ comments seem. Another study suggests that authors consider the competence of their reviewers to be closely aligned to the editorial decision [@drvenica2019peer]. Just be aware that this is normal. Remember that the reviewers are humans, and they have sat down and given freely of their own time to read your work. The most important thing to be aware of is that all they had was what you had written. No background information, and possibly no information about the species or the system involved. They will be experts at some level, but perhaps not the type you might expect. Importantly, the editor asked them because they thought that their opinion would be of importance in helping them make their decision on your paper. This means that you also need to respect their opinion and comments, even if you don’t agree with them or find them to be offensive, arrogant or even rude. Remember also that some apparent rudeness may just be a reviewer who has a sense of humour that you don’t understand. There are lots of examples of this at [ShitMyReviewersSay](https://shitmyreviewerssay.tumblr.com/). So no matter what you think of each comment, you should respond to it in a professional and courteous manner that shows that you are a professional scientist.

Why do scientists make disparaging or unprofessional remarks to their colleagues in peer review? Whenever two or three scientists get together, you hear tales of recent woes associated with peer review. The retelling of such stories is all part of the collective, cathartic unburdening of what can be a traumatic experience especially when we put so much effort into each piece of work [see @hyland2020this]. Reading through a lot of these reviewers’ comments, I can see that there is an attempt at humour. This humour is not appreciated by those who receive the reviews. Perhaps I understand the humour, because I also come from that same culture that dominates STEM, but that is not understood or even recognised as humour by others. Writing humorous reviews is unprofessional, especially if it is used to accentuate negative aspects. Needless to say, we could all do without unprofessional reviews.

## _Ad hominem_ attacks

One of the shocking results of a very large study of peer review of PLOS ONE articles is the large number of comments that are written directly attacking the authors as a group or personally [i.e. _ad hominem_ attacks, see @eve2021reading]. This should not happen. Reviewers should be confining their objective comments to the work and its presentation. However, this is an aspect of peer review where authors (especially the corresponding and leading authors) will need to acquire a thick skin, because unprofessional comments are made to people across gender and racial groupings, but especially toward traditionally underrepresented groups [@silbiger2019unprofessional]. Sadly, these same groups feel that such comments disproportionately impact their productivity and career advancement [@silbiger2019unprofessional]. Reading comments that are sent to other authors can be cathartic as these allow you to see that everyone receives such negative comments. [ShitMyReviewers](https://shitmyreviewerssay.tumblr.com/) is a good source of these, or see Eve et al [-@eve2021reading], or Silbiger and Stubler [-@silbiger2019unprofessional]. When _ad hominem_ attacks are made, it would be good if editors openly and explicitly identified these as **bad behaviour**. It would certainly improve the understanding of authors if editors intervened when such _ad hominem_ attacks are made. This would not necessarily involve deleting these comments, but directing authors to ignore the same.

**Why do academics make all of these terrible comments?** I can't pretend to know the answer for all of the cases, but I can speak from personal experience that time is at a premium, and time spent reading and reviewing manuscripts tends to be quality time. If these manuscripts are not of a quality that will pass peer review (i.e. will be rejected), then this feels like an abuse of professional time - especially when editors should have spotted the same mistake in their [first reading](editors.html). Editors that fail to see manuscripts that [should be rejected](#desk) do the reviewers a dis-service by increasing the amount of work for everyone (more people and more time is involved). Resentment and frustration may follow on the part of reviewers that manifests itself in the form of _ad hominem_ attacks.

## Demonstrated biases in peer review

Although Table \@ref(tab:Bias) shows that many kinds of bias have been explicitly demonstrated, that’s certainly not their limit. Given that over 280 biases have already been catalogued (I encourage you to look through the [online catalogue](www.catalogofbias.org), many more different types of bias are likely to exist in peer review. Let’s not forget that our biases have evolved because they are very useful. They exist as a way of shortcutting exhaustive decision making based on random variables. But maybe peer review needs some more of this. And perhaps that means that I should be tolerant when I’m asked to review an economics journal, as these folk clearly weren’t exhibiting any biases associated with economists when they picked me (see [below](editors.html)).

Table: (\#tab:Bias) **There are as many biases in peer review as there are humans that conduct them.** This table demonstrates some of the biases that have been proven in studies.

| Bias for which there is evidence | Study demonstrating bias | 
| - | - | 
| Against female authors | @tregenza2002gender; @manlove2018authors; @fox2019gender; @budden2008double-blind; @morgan2018foundation |
| Against female reviewers | @helmer2017gender; @fox2019gendera |
| Towards author reputation, favouring acceptance of manuscripts despite poor reviews | @bravo2018hidden; @okike2016single-blind |
| Towards authors from more prestigious institutions, also called prestige bias | @ceci1982peer; @travis1991new; @tomkins2017single; @manlove2018authors ; @lee2013bias |
| Nationality and language bias | @lee2013bias; @manlove2018authors; @nunez2021monolingual; @link1998us |
| Confirmation bias (the tendency for journals and reviewers to favour significant results) | @mahoney1977publication; @fanelli2010positive; @fanelli2012negative  [Part I](transparency.html) |
| Publication bias (the literature contains a bias in published results) | @munafo2007association; @dongen2011associations; @fanelli2017meta-assessment; @franco2014publication [Part IV](#badscience) |

Perhaps the biggest problem facing those who wish to reform the peer review system is that it all starts with editors who are choosing reviewers. Those editors themselves have their own inherent biases. When they look for reviewers, they are likely to sample from within their own group of peers who have the same biases. Interestingly, bias (in general) is more easily perceived by early career scientists [@zvereva2021biases]. My experience is that soliciting reviews from people that I don’t know and have no connection with (are outside of my field) are more likely to fail - they will say no, or they won’t reply to the request [see @perry2012peer]. This is even for academics that are publishing within the same area.

@@

## Decisions rest with editors
A good editor will look at the reasoning in the reviews and make a decision in an unbiased way. A poor editor may be swayed by the perceived influence of an important reviewer irrespective of their argument. An increasing trend that I’ve noticed is that editors will simply take a decision that follows the consensus of all reviewers: that is, they rate all reviewers equally [see also @rothwell2000reproducibility]. However, I would argue that this is also bad editing. Irrespective of the bias from reviewers, guarding the integrity of the process of peer review lies with editors.

@@

Casnici et al [-@casnici2017assessing] tracked the fate of rejected manuscripts and showed that if the reviewers had several rounds of peer review before rejection, these manuscripts benefited later by being accepted to journals with a higher Impact Factor, and/or obtaining greater numbers of citations, even if the reviewers were instrumental in rejecting the manuscript. This suggests that in working collaboratively on a manuscript, reviewers are more likely to promote, cite and help authors. The alternative is that reviewers agree to re-review an article again because they see merit in it, even if they also see flaws. And having spent considerable effort on manuscript, they are more likely to remember and cite it. But this doesn’t take away from the idea that reviewers and authors are collaborating in a social way.

## Fixing peer review
Fixing peer review will rest with the community of biological sciences, at the level of editors and the scholarly societies that they represent. To me, it is clear that we won't fix peer review by asking our peers to be less biased, or by asking them to be more rational. We should know by now that we can't fix people in this way. For example, Khoo [-@khoo2018there] found that there was little improvement in reviews after [reviewer training courses](#onlinecourse), even when these included feedback on previous reviews submitted. Instead, we have to plot a course for peer review whereby we accept that reviews will contain bias and irrational content, and train those in editorial positions to try to spot these, instead of falling victim to them.

There are lots of ways in which to do improve editorial oversight. My intuition is that the crux is to find a way that makes it more efficient and objective for the editors. For example, to try to pin down reviewers on where they find fault and exactly what that fault is. There is a difference between:

 - insufficient information to decide whether the experimental design was faulty 
  + and needing this clarified before a decision can be made.
 - and finding a fundamental error in the experimental design such that the manuscript can be rejected
 - and insufficient power (in replicates or sampling) to reach the conclusion generated in the manuscript.

A manuscript having each of these outcomes should have different fates: The first is Reject & Resubmit, the second is Reject, while the third may warrant either major/minor revisions (depending what else is problematic), or movement to another journal.

However, although the minority of peer reviewers means that at times prejudices and biases will align, a more inclusive world might mean that they diverge, prompting more differences in opinions about what should happen to manuscripts. Given that it is already quite difficult to find enough reviewers, simply asking more reviewers won't fix this. Instead, we need ways in which editors can more easily come to decisions on  manuscripts taking into consideration the potential faults. This really entails journals being more transparent about what flaws in manuscripts will be considered fatal. For journals where methodological competency is all that's required (see here for [commiment to publish](#commitment)), this is simple, but for many more journals (particularly those that are important for Early Career Researchers because advancement in their careers will depend on publishing there), there will be more ill defined, editor-centric choices that are more about fashion in Biological Sciences, than good science _per se_.

Hence, fixing peer review comes back to fixing problems associated with the publishing culture (and all that that entails), rather than any a simple fix-all for the myriad of existing publishing options. Hence we have a book on publishing for early career researchers!
